---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
  word_document: default
---


## Loading and preprocessing the data
```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(caret)
library(RANN)
library(gridExtra)
```

```{r}
data <- read.csv("activity.csv")
data$date <- as.Date(data$date, "%Y-%m-%d") ##parse dates
data$weekday <- weekdays(data$date) ## gets the weekday
data$weekend <- data$weekday %in% c("Saturday", "Sunday") ## flag for Weekwend/weekeday. TRUE =  = weeend
```
## What is mean total number of steps taken per day?

```{r}
groupedByDay <- data %>% group_by(date) %>% summarize (total = sum(steps, na.rm =TRUE))
hist1<-ggplot(groupedByDay, aes(total)) + geom_histogram()
hist1
meanByTotalNumberOfStepsPerDay <- mean(groupedByDay$total, na.rm =TRUE)
medianByTotalNumberOfStepsPerDay <- median(groupedByDay$total, na.rm =TRUE) 
meanByTotalNumberOfStepsPerDay
medianByTotalNumberOfStepsPerDay
```
## What is the average daily activity pattern?
```{r}
byInterlval <-data %>% group_by(interval) %>% summarize(avg = mean(steps, na.rm=TRUE)) %>% arrange(desc(avg))
ggplot(byInterlval, aes(interval, avg)) + geom_line() +
   xlab("interval") + ylab("average steps")


FiveMinIntervalWithTheMaxNumberOfSteps <- byInterlval[1,]
FiveMinIntervalWithTheMaxNumberOfSteps
```
## Imputing missing values
```{r}
## Median imputation through  the caret package
pp<- preProcess(data, method="medianImpute") 
medianImpute <-predict(pp, data)$steps

## new dataset with missing data filled in
dataMedianImpute <- data
dataMedianImpute$steps <- medianImpute

##Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
groupedByDayImpute <- dataMedianImpute %>% group_by(date) %>% summarize (total = sum(steps, na.rm =TRUE))
hist2<-ggplot(groupedByDayImpute, aes(total)) + geom_histogram()
grid.arrange(hist1, hist2, ncol=2)


meanByTotalNumberOfStepsPerDayImpute <- mean(groupedByDayImpute$total, na.rm =TRUE)
medianByTotalNumberOfStepsPerDayImpute <- median(groupedByDayImpute$total, na.rm =TRUE) 
meanByTotalNumberOfStepsPerDayImpute
medianByTotalNumberOfStepsPerDayImpute
```

Impact of imputing the data is almost non existent for this dataset. Neither the summary statistics nor the histogram show any relevant impact.



## Are there differences in activity patterns between weekdays and weekends?

```{r}

## weekeday/weekend column was calculated in the Loading and preprocessing the data section
byInterlvalWeekend <-data %>% group_by(interval,weekend) %>% summarize(avg = mean(steps, na.rm=TRUE)) %>% arrange(desc(avg))
line1<- ggplot(byInterlvalWeekend %>% filter(weekend), aes(interval, avg)) + geom_line() + ggtitle("Weekend") +
   xlab("interval") + ylab("average steps") + geom_vline(xintercept=c(1000,2200), linetype="dotted", colour = "red", size= 1)
line2<- ggplot(byInterlvalWeekend %>% filter(!weekend), aes(interval, avg)) + geom_line() + ggtitle("Weekday") +
   xlab("interval") + ylab("average steps") + geom_vline(xintercept=c(1000,2200), linetype="dotted", colour = "red", size= 1)
grid.arrange(line1, line2, nrow=2)

```

The steps taken on weekends between on intervals from 1000-2200 are higher than those on weekdays. See the section on both histograms between the vertical red lines.